# Demographics and Employment in the US
##### *Unit-1: Introduction to Analytics*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```
</br>

In the wake of the Great Recession of 2009, there has been a good deal of focus 
on employment statistics, one of the most important metrics policymakers use to 
gauge the overall strength of the economy. In the United States, the government 
measures unemployment using the Current Population Survey (CPS), which collects 
demographic and employment information from a wide range of Americans each 
month. In this exercise, we will employ the topics reviewed in the lectures as 
well as a few new techniques using the September 2013 version of this rich, 
nationally representative dataset (available 
[online](http://thedataweb.rm.census.gov/ftp/cps_ftp.html)).

The observations in the dataset represent people surveyed in the September 2013 
CPS who actually completed a survey. While the full dataset has 385 variables, 
in this exercise, we will use a more compact version of the dataset, 
[CPSData.csv](https://courses.edx.org/c4x/MITx/15.071x_2/asset/CPSData.csv), 
which has the following variables:

- `PeopleInHousehold`: The number of people in the interviewee's household.

- `Region`: The census region where the interviewee lives.

- `State`: The state where the interviewee lives.

- `MetroAreaCode`: A code that identifies the metropolitan area in which the 
interviewee lives (missing if the interviewee does not live in a metropolitan 
area). The mapping from codes to names of metropolitan areas is provided in the 
file [MetroAreaCodes.csv](https://courses.edx.org/c4x/MITx/15.071x_2/asset/MetroAreaCodes.csv).

- `Age`: The age, in years, of the interviewee. 80 represents people aged 
80-84, and 85 represents people aged 85 and higher.

- `Married`: The marriage status of the interviewee.

- `Sex`: The sex of the interviewee.

- `Education`: The maximum level of education obtained by the interviewee.

- `Race`: The race of the interviewee.

- `Hispanic`: Whether the interviewee is of Hispanic ethnicity.

- `CountryOfBirthCode`: A code identifying the country of birth of the 
interviewee. The mapping from codes to names of countries is provided in the 
file [CountryCodes.csv](https://courses.edx.org/c4x/MITx/15.071x_2/asset/CountryCodes.csv).

- `Citizenship`: The United States citizenship status of the interviewee.

- `EmploymentStatus`: The status of employment of the interviewee.

- `Industry`: The industry of employment of the interviewee (only available if 
they are employed).

----------

#### Summary Statistics
Load the dataset from [CPSData.csv](https://courses.edx.org/c4x/MITx/15.071x_2/asset/CPSData.csv)
into a data frame called CPS, and view the dataset with `summary()` and `str()` 
commands.

```{r}
CPS = read.csv("CPSData.csv")
```
</br>

##### 1.1 How many interviewees are there in the dataset?  
```{r}
nrow(CPS)
```
</br>

##### 1.2 Among the interviewees with a value reported for the Industry variable, what is the most common industry of employment?
```{r}
which.max(table(CPS$Industry))
```
</br>

##### 1.3 Which state has the fewest and the largest number of interviewees?
Recall from the homework assignment "The Analytical Detective" that you can 
call the `sort()` function on the output of the `table()` function to obtain a 
sorted breakdown of a variable. For instance, `sort(table(CPS$Region))` sorts the 
regions by the number of interviewees from that region.

**Which state has the fewest interviewees?**  
```{r}
sort(table(CPS$State))[1]
```

**Which state has the largest number of interviewees?**  
```{r}
sort(table(CPS$State), decreasing=TRUE)[1]
```
</br>

##### 1.4 What proportion of interviewees are citizens of the United States?
```{r}
table(CPS$Citizenship)
```
(123712/131302)=0.942  
</br>

##### 1.5 For which races are there at least 250 interviewees are of Hispanic enthicity?
The CPS differentiates between race (with possible values American Indian, 
Asian, Black, Pacific Islander, White, or Multiracial) and ethnicity. A number 
of interviewees are of Hispanic ethnicity, as captured by the Hispanic variable.

```{r}
table(CPS$Race, CPS$Hispanic)
```
American Indian, Black, Multiracial and White.  
</br>

#### Evaluating Missing Values

##### 2.1 Which variables have at least one interview with a missing (NA) value?
```{r}
summary(CPS)
```
The relevant variables can be noted from the above output.
</br>

##### 2.2 Which is the most accurate regarding missing values?
Often when evaluating a new dataset, we try to identify if there is a pattern 
in the missing values in the dataset. We will try to determine if there is a 
pattern in the missing values of the Married variable. The function 
`is.na(CPS$Married)` returns a vector of TRUE/FALSE values for whether the 
Married variable is missing. We can see the breakdown of whether Married is 
missing based on the reported value of the Region variable with the function 
`table(CPS$Region, is.na(CPS$Married)`).

```{r}
table(CPS$Region, is.na(CPS$Married))
table(CPS$Sex, is.na(CPS$Married))
table(CPS$Age, is.na(CPS$Married))[0:20,]
table(CPS$Citizenship, is.na(CPS$Married))
```
The Married variable being missing is related to the Age value for the 
interviewee. For each possible value of Region, Sex, and Citizenship, there are
both interviewees Aged 0-14 and is present for all interviewees aged 15 and 
older. This is because the CPS does not ask about marriage status for 
interviewees 14 or younger.  
</br>

##### 2.3 How many states had all interviewees living in a non-metropolitan area?
As mentioned in the variable descriptions, MetroAreaCode is missing if an 
interviewee does not live in a metropolitan area. Using the same technique as 
in the previous question, answer the following questions about people who live 
in non-metropolitan areas.  

**How many states had all interviewees living in a non-metropolitan area** 
**(aka they have a missing MetroAreaCode value)? For this question, treat the** 
**District of Columbia as a state (even though it is not technically a state).**

```{r}
table(CPS$State, is.na(CPS$MetroAreaCode))
sum(table(CPS$State, is.na(CPS$MetroAreaCode))[,'FALSE'] == 0)
```

**How many states had all interviewees living in a metropolitan area? Again,** 
**treat the District of Columbia as a state.**

```{r}
sum(table(CPS$State, is.na(CPS$MetroAreaCode))[,'TRUE'] == 0)
```
</br>

##### 2.4 Which region of the United States has the largest proportion of interviewees living in a non-metropolitan area?

```{r}
table(CPS$Region, is.na(CPS$MetroAreaCode))
```
Midwest.  
</br>

##### 2.5 Proportion of interviewees.
While we were able to use the `table()` command to compute the proportion of 
interviewees from each region not living in a metropolitan area, it was 
somewhat tedious (it involved manually computing the proportion for each 
region) and isn't something you would want to do if there were a larger number 
of options. It turns out there is a less tedious way to compute the proportion 
of values that are TRUE. The `mean()` function, which takes the average of the 
values passed to it, will treat TRUE as 1 and FALSE as 0, meaning it returns 
the proportion of values that are true. For instance, 
`mean(c(TRUE, FALSE, TRUE, TRUE))` returns 0.75. Knowing this, use `tapply()` 
with the mean function to answer the following questions:  

**Which state has a proportion of interviewees living in a non-metropolitan**
**area closest to 30%?**
```{r}
sort(tapply(is.na(CPS$MetroAreaCode), CPS$State, mean))
```
From this output, it is noted that Wisconsin is the state closest to having 30%
of its interviewees from a non-metropolitan area (with 29.933% non-metropolitan
interviewees) and Montana is the state with the highest proportion of 
non-metropolitan interviewees without them all being non-metropolitan, at 83.6%.

**Which state has the largest proportion of non-metropolitan interviewees,** 
**ignoring states where all interviewees were non-metropolitan?**
Montana  
</br>

#### Integrating Metropolitan Area Data
Codes like MetroAreaCode and CountryOfBirthCode are a compact way to encode 
factor variables with text as their possible values and they are therefore 
quite common in survey datasets. In fact, all but one of the variables in this 
dataset were actually stored by a numeric code in the original CPS datafile.

When analyzing a variable stored by a numeric code, we will often want to 
convert it into the values the codes represent. To do this, we will use a 
dictionary, which maps the code to the actual value of the variable. We 
have provided dictionaries [MetroAreaCodes.csv](https://courses.edx.org/c4x/MITx/15.071x_2/asset/MetroAreaCodes.csv) 
and [CountryCodes.csv](https://courses.edx.org/c4x/MITx/15.071x_2/asset/CountryCodes.csv), 
which respectively map MetroAreaCode and CountryOfBirthCode into their true 
values. Read these two dictionaries into data frames MetroAreaMap and CountryMap.

```{r}
MetroAreaMap <- read.csv("MetroAreaCodes.csv")
CountryMap <- read.csv("CountryCodes.csv")
```
</br>

##### 3.1 What are the number of observations in both the data frames?
```{r}
nrow(MetroAreaMap)
nrow(CountryMap)
```
</br>

##### 3.2 Updating the CPS Data Frame.
To merge in the metropolitan areas, we want to connect the field MetroAreaCode 
from the CPS data frame with the field Code in MetroAreaMap. The following 
command merges the two data frames on these columns, overwriting the CPS data 
frame with the result:

```{r}
CPS = merge(CPS, MetroAreaMap, by.x="MetroAreaCode", by.y="Code", all.x=TRUE)
```

The first two arguments determine the data frames to be merged (they are called 
"x" and "y", respectively, in the subsequent parameters to the merge function). 
`by.x="MetroAreaCode"` means we're matching on the `MetroAreaCode` variable 
from the "x" data frame (CPS), while `by.y="Code"` means we're matching on the 
Code variable from the "y" data frame (MetroAreaMap). Finally, `all.x=TRUE` 
means we want to keep all rows from the "x" data frame (CPS), even if some of 
the rows' MetroAreaCode doesn't match any codes in MetroAreaMap (for those 
familiar with database terminology, this parameter makes the operation a left 
outer join instead of an inner join).

**Review the new version of the CPS data frame with the `summary()` and**
**`str()` functions. What is the name of the variable that was added to the** 
**data frame by the `merge()` operation?**

```{r}
str(CPS)
```
`MetroArea` was added.  
</br>

**How many interviewees have a missing value for the new metropolitan area**
**variable? Note that all of these interviewees would have been removed from**
**the merged data frame if we did not include the all.x=TRUE parameter.**

```{r}
table(is.na(CPS$MetroArea))['TRUE']
```
</br>

##### 3.3 Which metropolitan area has the largest number of interviewees?
```{r}
sort(table(CPS$MetroArea), decreasing = TRUE)[1:8] 
```
</br>

##### 3.4 Which metro-area has the highest proportion of interviewees of Hispanic ethnicity?
```{r}
sort(tapply(CPS$Hispanic, CPS$MetroArea, mean), decreasing = TRUE)[1:3]
```
It is noted that 96.6% interviewees from Laredo, TX, are from Hispanic 
ethnicity, the highest proportion among metro areas in the United States.  
</br>

##### 3.5 Determine the number of metro areas in the US from which at least 20% interviewees are Asian.
```{r}
sort(tapply(CPS$Race == "Asian", CPS$MetroArea, mean), decreasing = TRUE)[1:6]
```
We can read from the sorted output that Honolulu, HI; San-Francisco-Oakland-Fremont,
CA; San Jose-Sunnyvale-Santa Clara, CA; and Vallejo-Fairfield, CA had at least
20% of their interviewees of the Asian race.  
</br>

##### 3.6 Calculating Amidst Missing Data
Normally, we would look at the sorted proportion of interviewees from each 
metropolitan area who have not received a high school diploma with the command:
`sort(tapply(CPS$Education == "No high school diploma", CPS$MetroArea, mean))`

However, none of the interviewees aged 14 and younger have an education value 
reported, so the mean value is reported as NA for each metropolitan area. To 
get mean (and related functions, like sum) to ignore missing values, you can 
pass the parameter na.rm=TRUE. Passing na.rm=TRUE to the tapply function, 
determine which metropolitan area has the smallest proportion of interviewees 
who have received no high school diploma.

```{r}
sort(tapply(CPS$Education == "No high school diploma", CPS$MetroArea, mean, na.rm=TRUE))[1]
```
We cam see that Iowa City, IA had 2.9% of interviewees not finish high school,
the smallest value of any metropolitan area.  
</br>

#### Integrating Country of Birth Data
Just as we did with the metropolitan area information, merge in the country of 
birth information from the CountryMap data frame, replacing the CPS data frame 
with the result. If you accidentally overwrite CPS with the wrong values, 
remember that you can restore it by re-loading the data frame from CPSData.csv 
and then merging in the metropolitan area information using the command 
provided in the previous subproblem.

```{r}
CPS <- merge(CPS, CountryMap, by.x="CountryOfBirthCode", by.y="Code", all.x=TRUE)
```
</br>

##### 4.1 Answer the following questions on the new updated Data Frame.
**What is the name of the variable added to the CPS data frame by this**
**merge operation?** 
```{r}
str(CPS)
```
Country  
</br>

**How many interviewees have a missing value for the new country of birth** 
**variable?** 
```{r}
summary(CPS)
```
The answer can be noted from the above output.
</br>

##### 4.2 Among all interviewees born outside of North America, which country was the most common place of birth?
```{r}
sort(table(CPS$Country), decreasing=TRUE)[1:3]
```
</br>

##### 4.3 What proportion of interviewees from the NY-NJ-PA metropolitan area have a country of birth that is not the US? For this computation, don't include people with missing data.
```{r}
table(CPS$MetroArea == "New York-Northern New Jersey-Long Island, NY-NJ-PA", CPS$Country != "United States")
```
It is seen that 1668 of interviewees from this metropolitan area were born 
outside the United States and 2736 were born in the United States (it turns out
an additional 5 have a missing country of origin). Therefore, the proportion is
1668/(1668+3736)=0.309.  
</br>

##### 4.4 Which metropolitan area has the largest number of interviewees with a country of birth ____?
**in India?**  
```{r}
sort(tapply(CPS$Country=="India", CPS$MetroArea, sum, na.rm=T), decreasing = TRUE)[1]
```
**in Brazil?**  
```{r}
sort(tapply(CPS$Country=="Brazil", CPS$MetroArea, sum, na.rm=T), decreasing = TRUE)[1]
```
**in Somalia?**  
```{r}
sort(tapply(CPS$Country=="Somalia", CPS$MetroArea, sum, na.rm=T), decreasing = TRUE)[1]
```
</br>

*End of Homework Assignment*