# An Analytical Detective
##### *Unit-1: Introduction to Analytics*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```
</br>

Crime is an international concern, but it is documented and handled in very 
different ways in different countries. In the United States, violent crimes and
property crimes are recorded by the Federal Bureau of Investigation (FBI).
Additionally, each city documents crime and some cities release data regarding
crime rates. The city of Chicago, Illinois releases crime data from 2001 onward
[online](https://goo.gl/N1fQfF).

Chicago is the third most populous city in the United States, with a population 
of over 2.7 million people. 

There are two main types of crimes: violent crimes, and property crimes. In 
this problem, we'll focus on one specific type of property crime, called "motor 
vehicle theft" (sometimes referred to as grand theft auto). This is the act of 
stealing, or attempting to steal, a car. In this problem, we'll use some basic 
data analysis in R to understand the motor vehicle thefts in Chicago.

Please download the file [mvtWeek1.csv](http://bit.ly/2p4175k) for this problem 
(do not open this file in any spreadsheet software before completing this 
problem because it might change the format of the Date field). Here is a list 
of descriptions of variables:

- `ID`: a unique identifier for each observation.

- `Date`: the date the crime occurred.

- `LocationDescription`: the location where the crime occurred.

- `Arrest`: whether or not an arrest was made for the crime (TRUE if an 
arrest was made, and FALSE if an arrest was not made)

- `Domestic`: whether or not the crime was a domestic crime, meaning that it 
was committed against a family member (TRUE if it was domestic, and FALSE if it 
was not domestic)

- `Beat`: the area, or "beat" in which the crime occurred. This is the 
smallest regional division defined by the Chicago police department.

- `District`: the police district in which the crime occurred. Each district 
is composed of many beats, and are defined by the Chicago Police Department.

- `CommunityArea`: the community area in which the crime occurred. Since the 
1920s, Chicago has been divided into what are called "community areas," of 
which there are now 77. The community areas were devised in an attempt to 
create socially homogeneous regions.

- `Year`: the year in which the crime occurred.

- `Latitude`: the latitude of the location at which the crime occurred.

- `Longitude`: the longitude of the location at which the crime occurred.

----------

#### Loading the Data
Read the dataset [mvtWeek1.csv](http://bit.ly/2p4175k) into R, using the 
`read.csv` function, and call the data frame "mvt". Remember to navigate to the 
directory on your computer containing the file `mvtWeek1.csv` first. It may 
take a few minutes to read in the data, since it is pretty large. Then, use the
`str` and `summary` functions to answer the following questions.
  
```{R}
mvt = read.csv("mvtWeek1.csv")
```
</br>

##### 1.1 How many rows of data (observations) are in this dataset?
```{R}
str(mvt)
```
The first row of the output says that there are: 191641 observations.  
</br>
   
##### 1.2 How many variables are there in this dataset?  
The first row of the `str` output says that there are: 11 variables.  
</br>

##### 1.3 Using the "max" function, what is the maximum value of the variable "ID"?
```{R}
max(mvt$ID)
```
</br>

##### 1.4 What is the minimum value of the variable "Beat"?
```{R}
min(mvt$Beat)
```
</br>

##### 1.5 How many observations have value TRUE in the Arrest variable (number of crimes for which an arrest was made)?
```{R}
summary(mvt)
```
On inspecting the `summary` output, it can be seen that 15,536 observations 
fall under the category TRUE for the variable Arrest.  
</br>

##### 1.6 How many observations have value a LocationDescription value of ALLEY?
From the `summary` output, 2308 values fall under the category. Alternatively,
`table(mvt$LocationDescription)` can be used.  
</br>

#### Understanding Dates in R
In many datasets, like this one, you have a date field. Unfortunately, R does 
not automatically recognize entries that look like dates. We need to use a 
function in R to extract the date and time. Take a look at the first entry of 
Date (remember to use square brackets when looking at a certain entry of a 
variable).  
</br>

##### 2.1 In what format are the entries in the variable Date?
```{r}
mvt$Date[1]
```
The above output suggests that the date format is Month/Day/Year Hour:Minute.  
</br>

##### 2.2 Converting to Date object.
Now, let's convert these characters into a Date object in R.
```{r}
DateConvert = as.Date(strptime(mvt$Date, "%m/%d/%y %H:%M"))
```
This converts the variable "Date" into a Date object in R. Take a look at the 
variable DateConvert using the summary function.

**What is the month and year of the median date in our dataset?**   
```{r}
summary(DateConvert)
```
The median date is May 2006 (2006-05-21).  
</br>

##### 2.3 In which month did the fewest motor vehicle thefts occur?
Now, let's extract the month and the day of the week, and add these variables 
to our data frame mvt. We can do this with two simple functions. 
Type the following commands in R:
```{r}
mvt$Month = months(DateConvert)
mvt$Weekday = weekdays(DateConvert)
```

This creates two new variables in our data frame, Month and Weekday, and sets 
them equal to the month and weekday values that we can extract from the Date 
object. Lastly, replace the old Date variable with DateConvert by typing:

```{r}
mvt$Date = DateConvert
```

*Using the table command, answer the following questions.*

**In which month did the fewest motor vehicle thefts occur?**
```{r}
sort(table(mvt$Month))
```
</br>

##### 2.4 On which weekday did the most motor vehicle thefts occur?
```{r}
table(mvt$Weekday)
```
</br>

##### 2.5 Which month has the largest number of motor vehicle thefts for which an arrest was made?
Each observation in the dataset represent a motor vehicle theft, and the Arrest
variable indicates whether an arrest was later made on this theft. 
```{r}
table(mvt$Arrest, mvt$Month)
```
The largest number of observations with Arrest=TRUE are for January.  
</br>

#### Visualizing Crime Trends
Now, let's make some plots to help us better understand how crime has changed
over time in Chicago. Throughout this problem, and in general, you can save 
your plot to a file. For more information, [this website](http://bit.ly/2FDcrLQ)
very clearly explains the process. 

First, let's make a histogram of the variable Date. We'll add an extra
argument, to specify the number of bars we want in our histogram. In your R
console, type

```{r}
hist(mvt$Date, breaks=100)
```

*Looking at the histogram, answer the following questions.*

**Does it look like the crime increases or decreases from 2002-2012?**
Decreases.
</br>

**Does it look like the crime increases or decreases from 2005-2008?**
Decreases.
</br>

**Does it look like the crime increases or decreases from 2009-2011?**
Increases.  
</br>

##### 3.2 Using box-plots
Now, let's see how arrests have changed over time. Create a boxplot of the 
variable "Date," sorted by the variable "Arrest" (if you are not familiar with 
boxplots and would like to learn more, check out 
[this tutorial](http://msenux.redwoods.edu/math/R/boxplot.php)). In a boxplot, 
the bold horizontal line is the median value of the data, the box shows the 
range of values between the first quartile and third quartile, and the whiskers 
(the dotted lines extending outside the box) show the minimum and maximum 
values, excluding any outliers (which are plotted as circles). Outliers are 
defined by first computing the difference between the first and third quartile 
values, or the height of the box. This number is called the Inter-Quartile 
Range (IQR). Any point that is greater than the third quartile plus the IQR or 
less than the first quartile minus the IQR is considered an outlier.

**Does it look like there were more crimes for which arrests were made in the** 
**first half of the time period or the second half of the time period?**
(Note that the time period is from 2001 to 2012, so the middle of the 
time period is the beginning of 2007.)

```{r}
boxplot(mvt$Date ~ mvt$Arrest)
```

From the boxplot, it can be seen that for Arrest=TRUE, it is skewed towards
the bottom of the plot, meaning that there were more crimes for which arrests
were made in the first half of the time period.  
</br>

##### 3.3 For what proportion of motor vehicle thefts in 2001 was an arrest made?
Note: in this question and many others in the course, we are asking for an 
answer as a proportion. Therefore, your answer should take a value between 
0 and 1.

```{r}
table(mvt$Arrest, mvt$Year)
```
The required proportion is 2152/(2152+18157) = 0.1041173  
</br>

##### 3.4 For what proportion of motor vehicle thefts in 2007 was an arrest made?
The required proportion is 1212/(1212+13068) = 0.08487395  
</br>

##### 3.5 For what proportion of motor vehicle thefts in 2012 was an arrest made?
The required proportion is 550/(550+13542) = 0.03902924  
</br>

Since there may still be open investigations for recent crimes, this could 
explain the trend we are seeing in the data. There could also be other factors 
at play, and this trend should be investigated further. However, since we don't 
know when the arrests were actually made, our detective work in this area has 
reached a dead end.  
</br>

#### Popular Locations
Analyzing this data could be useful to the Chicago Police Department when 
deciding where to allocate resources. If they want to increase the number of 
arrests that are made for motor vehicle thefts, where should they focus their 
efforts?

We want to find the top five locations where motor vehicle thefts occur. If you 
create a table of the LocationDescription variable, it is unfortunately very 
hard to read since there are 78 different locations in the data set. By using 
the sort function, we can view this same table, but sorted by the number of 
observations in each category. In your R console, type: 
`sort(table(mvt$LocationDescription))`  
</br>

##### 4.1 Which locations are the top five locations for motor vehcle thefts?
```{r}
sort(table(mvt$LocationDescription), decreasing = TRUE)[1:6]
```
</br>

##### 4.2 Analyse Top-5 Locations
Create a subset of your data, only taking observations for which the theft 
happened in one of these five locations, and call this new data set "Top5". 
To do this, you can use the | symbol. In the lecture, we used the & symbol to 
use two criteria to make a subset of the data. To only take observations that 
have a certain value in one variable or the other, the | character can be used 
in place of the & symbol. This is also called a logical "or" operation.

Alternately, you could create five different subsets, and then merge them 
together into one data frame using rbind.

**How many observations are in Top5?**
```{r}
Top5 = subset(mvt, LocationDescription=="STREET" |
                LocationDescription=="PARKING LOT/GARAGE(NON.RESID.)" | 
                LocationDescription=="ALLEY" | 
                LocationDescription=="GAS STATION" | 
                LocationDescription=="DRIVEWAY - RESIDENTIAL")

str(Top5)
```
</br>

##### 4.3 Which of these locations has a much higher arrest rate than the other locations?
R will remember the other categories of the LocationDescription variable from 
the original dataset, so running `table(Top5$LocationDescription)` will have a 
lot of unnecessary output. To make our tables a bit nicer to read, we can 
refresh this factor variable. In your R console, type:

```{r}
Top5$LocationDescription = factor(Top5$LocationDescription)
```

If you run the str function again on Top5, you should see that 
LocationDescription now only has 5 values, as we expect.

Use the Top5 data frame to answer the remaining questions.

```{r}
table(Top5$LocationDescription, Top5$Arrest)
```
On inspecting the percentage of arrests, it is found that *Gas Station* has the 
highest percentage of arrests, with over *20%* of motor vehicle thefts 
resulting in an arrest.  
</br>  

##### 4.4 On which day of the week do the most motor vehicle thefts at gas stations happen?
```{r}
table(Top5$LocationDescription, Top5$Weekday)
```
Saturday.  
</br>

##### 4.5 On which day of the week fo the fewest motor vehicle thefts in residential driveways happen?
```{r}
table(Top5$LocationDescription, Top5$Weekday)
```
Saturday.  
</br>  

*End of Homework Assignment*