# Internet Privacy Poll
##### *Unit-1: Introduction to Analytics*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```
</br>

Internet privacy has gained widespread attention in recent years. To measure 
the degree to which people are concerned about hot-button issues like Internet 
privacy, social scientists conduct polls in which they interview a large number 
of people about the topic. In this assignment, we will analyze data from a 
July 2013 Pew Internet and American Life Project poll on Internet anonymity and 
privacy, which involved interviews across the United States. While the full 
polling data can be found [here](http://bit.ly/2pmP5DZ), we will use a more 
limited version of the results, available in [AnonymityPoll.csv](https://courses.edx.org/c4x/MITx/15.071x_2/asset/AnonymityPoll.csv). 

The dataset has the following fields (all Internet use-related fields were only 
collected from interviewees who either use the Internet or have a smartphone):

- `Internet.Use`: A binary variable indicating if the interviewee uses the 
Internet, at least occasionally (equals 1 if the interviewee uses the Internet, 
and equals 0 if the interviewee does not use the Internet).


- `Smartphone`: A binary variable indicating if the interviewee has a 
smartphone (equals 1 if they do have a smartphone, and equals 0 if they don't 
have a smartphone).

- `Sex`: Male or Female.

- `Age`: Age in years.

- `State`: State of residence of the interviewee.

- `Region`: Census region of the interviewee (Midwest, Northeast, South, 
or West).

- `Conservativeness`: Self-described level of conservativeness of 
interviewee, from 1 (very liberal) to 5 (very conservative).

- `Info.On.Internet`: Number of the following items this interviewee believes 
to be available on the Internet for others to see: [1] Their email address; 
[2] Their home address; [3] Their home phone number; [4] Their cell phone 
number; [5] The employer/company they work for; [6] Their political party or 
political affiliation; [7] Things they've written that have their name on it; 
[8] A photo of them; [9] A video of them; [10] Which groups or organizations 
they belong to; [11] Their birth date.

- `Worry.About.Info`: A binary variable indicating if the interviewee worries 
about how much information is available about them on the Internet (equals 1 if 
they worry, and equals 0 if they don't worry).

- `Privacy.Importance`: A score from 0 (privacy is not too important) to 100 
(privacy is very important), which combines the degree to which they find 
privacy important in the following: [1] The websites they browse; [2] Knowledge 
of the place they are located when they use the Internet; [3] The content and 
files they download; [4] The times of day they are online; [5] The applications 
or programs they use; [6] The searches they perform; [7] The content of their 
email; [8] The people they exchange email with; [9] The content of their online 
chats or hangouts with others.

- `Anonymity.Possible`: A binary variable indicating if the interviewee thinks 
it's possible to use the Internet anonymously, meaning in such a way that 
online activities can't be traced back to them (equals 1 if he/she believes 
you can, and equals 0 if he/she believes you can't).

- `Tried.Masking.Identity`: A binary variable indicating if the interviewee has 
ever tried to mask his/her identity when using the Internet (equals 1 if he/she 
has tried to mask his/her identity, and equals 0 if he/she has not tried to 
mask his/her identity).

- `Privacy.Laws.Effective`: A binary variable indicating if the interviewee 
believes United States law provides reasonable privacy protection for Internet 
users (equals 1 if he/she believes it does, and equals 0 if he/she believes 
it doesn't).

----------

#### Loading and Summarizing the Dataset
Using read.csv(), load the dataset from 
[AnonymityPoll.csv](https://courses.edx.org/c4x/MITx/15.071x_2/asset/AnonymityPoll.csv) 
into a data frame called 'poll' and summarize it with the `summary()` and `str()`. 
</br>

##### 1.1 How many people participated in the poll?
```{r}
poll = read.csv("AnonymityPoll.csv")
str(poll)
summary(poll)
```
</br>

##### 1.2 Breakdown of the number of people with smartphones.
```{r}
table(poll$Smartphone)
summary(poll$Smartphone)
```

**How many interviewees responded that they use a smartphone?** 487 

**How many interviewees responded that they don't use a smartphone?** 472 

**How many interviewees did not respond to the question, resulting in a missing value?** 43  
</br>

##### 1.3 Use the table() function to answer the following questions:
By using the `table()` function on two variables, we can tell how they are 
related. To use the `table()` function on two variables, just put the two 
variable names inside the parentheses, separated by a comma. In the output, the
possible values of the first variable will be listed in the left, and the 
possible values of the second variable will be listed on the top. Each entry of 
the table counts the number of observations in the data set that have the value 
of the first value in that row, and the value of the second variable in that 
column. For example, suppose we want to create a table of the variables "Sex" 
and "Region". We would type

```{r}
table(poll$Sex, poll$Region)
```

This table tells us that we have 123 people in our dataset who are female and 
from the Midwest, 116 people in our dataset who are male and from the Midwest, 
90 people in our dataset who are female and from the Northeast, etc.
</br>

**Which of the following are states in the Midwest census region?**
```{r}
MidwestInterviewees = subset(poll, Region=="Midwest")
table(MidwestInterviewees$State)
```
</br>

**Which was the state in the South census region with the largest number of interviewees?**
```{r}
SouthInterviewees = subset(poll, Region=="South")
table(SouthInterviewees$State)
```
</br>

#### Internet and Smartphone Users
As mentioned in the introduction to this problem, many of the response 
variables (`Info.On.Internet`, `Worry.About.Info`, `Privacy.Importance`,
`Anonymity.Possible`, and `Tried.Masking.Identity`) were not collected if an
interviewee does not use the Internet or a smartphone, meaning the variables 
will have missing values for these interviewees.  
</br>

##### 2.1 Answer the following questions about the distribution of users.
**How many interviewees reported not having used the Internet and not having used a smartphone?**
```{r}
table(poll$Internet.Use, poll$Smartphone)
```
186
</br>

**How many interviewees reported having used the Internet and having used a smartphone?**
470
</br>

**How many interviewees reported having used the Internet but not having used a smartphone?**
285
</br>

**How many interviewees reported having used a smartphone but not having used the Internet?**
17  
</br>

##### 2.2 How many missing values?
```{r}
summary(poll)
```

**How many interviewees have a missing value for their Internet use?** 1
</br>

**How many interviewees have a missing value for their smartphone use?**
43  
</br>

##### 2.3 Subsetting the Data  
Use the subset function to obtain a data frame called "limited", which is 
limited to interviewees who reported Internet use or who reported smartphone 
use. In lecture, we used the & symbol to use two criteria to make a subset of 
the data. To only take observations that have a certain value in one variable 
or the other, the | character can be used in place of the & symbol. This is 
also called a logical "or" operation.

**How many interviewees are in the new data frame?**
```{r}
limited = subset(poll, Internet.Use==1 | Smartphone==1)
nrow(limited)
```
</br>

#### Summarizing Opinions About Internet Privacy

##### 3.1 Which variables have missing values in the limited data frame?
```{r}
summary(limited)
```
</br>

##### 3.2 What is the average number of pieces of personal information on the Internet?
```{r}
mean(limited$Info.On.Internet)
```
</br>

##### 3.3 - Summary of Info.On.Internet
**How many interviewees reported a value of 0 for Info.On.Internet?**
```{r}
table(limited$Info.On.Internet)
```
105
</br>

**How many interviewees reported the maximum value of 11 for Info.On.Internet?**
8  
</br>

##### 3.4 What proportion of interviewees who answered the Worry.About.Info question worry about how much information is available about them on the Internet? 

Note that to compute this proportion you will be dividing by the number of 
people who answered the Worry.About.Info question, not the total number of 
people in the data frame.

```{r}
sum(limited$Worry.About.Info==1, na.rm=T)/nrow(limited - 2)
# 2 is the number of NAs in this column
```
</br>

##### 3.5 What proportion of interviewees who answered the Anonymity.Possible question think it is possible to be completely anonymous on the Internet?

```{r}
sum(limited$Anonymity.Possible==1, na.rm=T)/(nrow(limited) - 39)
```
</br>

##### 3.6 What proportion of interviewees who answered the Tried.Masking.Identity question have tried masking their identity on the Internet?

```{r}
sum(limited$Tried.Masking.Identity==1, na.rm=T)/(nrow(limited) - 8)
```
</br>

##### 3.7 What proportion of interviewees who answered the Privacy.Laws.Effective question find the United States privacy laws effective?

```{r}
sum(limited$Privacy.Laws.Effective==1, na.rm=T)/(nrow(limited) - 65)
```
</br>

#### Relating Demographics to Polling Results
Often, we are interested in whether certain characteristics of interviewees 
(e.g., their age or political opinions) affect their opinions on the topic of 
the poll (in this case, opinions on privacy). In this section, we will 
investigate the relationship between the characteristics Age and Smartphone and 
outcome variables Info.On.Internet and Tried.Masking.Identity, again using the 
limited data frame we built in an earlier section of this problem.  
</br>

##### 4.1 Build a histogram of the age of interviewees. What is the best-represented age group in the population?

```{r}
hist(limited$Age)
```
</br>
People aged about 60 years old.  
</br>

##### 4.2 What is the largest number of interviewees that have exactly the same value in Age & Info.On.Internet variable?
Both Age and Info.On.Internet are variables that take on many values, so a good 
way to observe their relationship is through a graph. We learned in lecture 
that we can plot Age against Info.On.Internet with the command 
`plot(limited$Age, limited$Info.On.Internet)`. However, because Info.On.Internet 
takes on a small number of values, multiple points can be plotted in exactly 
the same location on this graph.

In other words, what is the largest number of overlapping points in the plot 
`plot(limited$Age, limited$Info.On.Internet)?`

```{r echo=FALSE, results='hide', message=FALSE} 
plot(limited$Age, limited$Info.On.Internet)
table(limited$Age, limited$Info.On.Internet)
```

```{r}
max(table(limited$Age, limited$Info.On.Internet))
```
</br>

##### 4.3 What appears to be the functionality of the jitter command based on the following information?
To avoid points covering each other up, we can use the jitter() function on the values we pass to the plot function. Experimenting with the command `jitter(c(1, 2, 3))`, what appears to be the functionality of the jitter command?

`jitter` adds or subtracts a small amount of random noise to the values passed 
to it, and two runs will yield different results.  
</br>

##### 4.4 Based on the plot, what relationship is observed between Age and Info.On.Internet?
```{r}
plot(jitter(limited$Age), jitter(limited$Info.On.Internet))
```

Older age seems moderately associated with a small value for Info.On.Internet. 
For younger people aged 18-30, the average value of Info.On.Internet appears to
be roughly 5, while most people aged 60 and older have value less than 5.  
</br>

##### 4.5 Using tapply() to obtain the summary of Info.On.Internet() broken down by smartphone user.

**What is the average Info.On.Internet value for smartphone users?**
```{r}
tapply(limited$Info.On.Internet, limited$Smartphone, summary)
```
4.368
</br>

**What is the average Info.On.Internet value for non-smartphone users?**
2.923  
</br>

##### 4.6 Using tapply() to break down Tried.Masking.Identity variable for smartphone/non-smartphone users.

**What proportion of smartphone users who answered the Tried.Masking.Identity** 
**question have tried masking their identity when using the Internet?**

```{r}
tapply(limited$Tried.Masking.Identity, limited$Smartphone, table)
```
Among smartphone users, 93 tried masking their identity and 390 did not,
resulting in proportion 93/(93+390)=0.1925.  
</br>

**What proportion of non-smartphone users who answered the** 
**Tried.Masking.Identity question have tried masking their identity when** 
**using the Internet?**

```{r}
tapply(limited$Tried.Masking.Identity, limited$Smartphone, mean, na.rm=T)
```

Next week, we will begin to more formally characterize how an outcome variable 
like Info.On.Internet can be predicted with a variable like Age or Smartphone.  
</br>

*End of Homework Assignment*